---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# prepR4pcm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Species names in your data rarely match the tip labels in your tree.
Formatting differences (`Homo_sapiens` vs `Homo sapiens`), taxonomic
synonyms, and simple typos silently drop species from phylogenetic
comparative analyses. **prepR4pcm** detects and resolves these mismatches
through a multi-stage matching cascade (exact, normalised, synonym, fuzzy),
documents every decision, and produces aligned data--tree pairs ready for
PGLS, phylogenetic mixed models, or any other PCM.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("itchyshin/prepR4pcm")
```

## Quick example

```{r example}
library(prepR4pcm)
library(ape)

# Reconcile a dataset against a phylogenetic tree
rec <- reconcile_tree(
  x         = avonet_subset,
  tree      = tree_jetz,
  x_species = "Species1",
  fuzzy     = TRUE,
  resolve   = "flag"
)
rec

# Apply the reconciliation: aligned data + pruned tree
aligned <- reconcile_apply(rec, data = avonet_subset, tree = tree_jetz,
                           species_col = "Species1", drop_unresolved = TRUE)
nrow(aligned$data)
ape::Ntip(aligned$tree)
```

## Features

- **Four-stage matching cascade**: exact match, normalised match (case,
  whitespace, underscores), synonym resolution via
  [taxadb](https://docs.ropensci.org/taxadb/) (Norman et al. 2020),
  and fuzzy matching with genus pre-filtering for typos.
- **Full provenance**: every name-matching decision is recorded.
  `reconcile_summary()`, `reconcile_plot()`, `reconcile_report()`, and
  `reconcile_suggest()` help you audit matches and find near-misses.
- **Multi-tree support**: `reconcile_to_trees()` matches a dataset against
  several trees at once; `reconcile_diff()` compares results.
- **Crosswalks and overrides**: `reconcile_crosswalk()` converts published
  taxonomy crosswalks (e.g., BirdLife--BirdTree) into override tables.
- **Tree augmentation**: `reconcile_augment()` grafts unresolved species
  onto a tree using genus-level placement (always run sensitivity
  analyses with and without augmented tips).

## Typical workflow

```
Load data + tree
       |
  reconcile_tree()
       |
  Review: reconcile_summary(), reconcile_plot(), reconcile_report()
       |
  Fix: reconcile_override(), reconcile_suggest()
       |
  reconcile_apply()
       |
  Aligned data + pruned tree --> PGLS / PGLMM
```

## Vignettes

- [Getting Started](https://itchyshin.github.io/prepR4pcm/docs/articles/getting-started.html)
  ([source](vignettes/getting-started.Rmd))
  --- core concepts and a minimal worked example
- [Bird Trait Workflow](https://itchyshin.github.io/prepR4pcm/docs/articles/bird-workflow.html)
  ([source](vignettes/bird-workflow.Rmd))
  --- a realistic multi-dataset, multi-tree analysis pipeline

## Citation

If you use prepR4pcm in your research, please cite the package:

```r
citation("prepR4pcm")
```

## Key dependencies

- [ape](https://cran.r-project.org/package=ape) --- phylogenetic tree
  handling (Paradis & Schliep 2019, *Bioinformatics* 35:526--528)
- [taxadb](https://docs.ropensci.org/taxadb/) --- local taxonomic
  synonym resolution (Norman et al. 2020, *Methods in Ecology and
  Evolution* 11:1153--1159)

## Bundled data sources

The package includes subset datasets for examples and testing. Full
credit to the original data providers:

- **AVONET**: Tobias et al. (2022) *Ecology Letters* 25:581--597.
  [doi:10.1111/ele.13898](https://doi.org/10.1111/ele.13898)
- **NestTrait v2**: Chia et al. (2023) *Scientific Data* 10:444.
  [doi:10.1038/s41597-023-02360-9](https://doi.org/10.1038/s41597-023-02360-9)
- **Plumage lightness**: Delhey et al. (2019) *American Naturalist*
  194:13--27. [doi:10.1086/703588](https://doi.org/10.1086/703588)
- **Jetz phylogeny**: Jetz et al. (2012) *Nature* 491:444--448.
  [doi:10.1038/nature11631](https://doi.org/10.1038/nature11631)
- **Clements checklist**: Clements et al. (2025) eBird/Clements
  Checklist of Birds of the World.
- **BirdLife-BirdTree crosswalk**: Tobias et al. (2022).

## License

MIT
