---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# prepR4pcm

<!-- badges: start -->
[![R-CMD-check](https://github.com/itchyshin/prepR4pcm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/itchyshin/prepR4pcm/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**prepR4pcm** reconciles species names across datasets and phylogenetic trees
for comparative biology workflows. Taxonomic mismatches --- formatting
differences, synonymy, and typos --- are a persistent source of silent data
loss in phylogenetic comparative methods (PCM). This package detects and
resolves those mismatches through a multi-stage matching cascade, producing
documented, reproducible alignments between data and trees.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("itchyshin/prepR4pcm")
```

## Quick example

```{r example}
library(prepR4pcm)
library(ape)

# Reconcile a dataset against a phylogenetic tree
rec <- reconcile_tree(
  x         = avonet_subset,
  tree      = tree_jetz,
  x_species = "Species1",
  fuzzy     = TRUE,
  resolve   = "flag"
)
rec

# Apply the reconciliation: aligned data + pruned tree
aligned <- reconcile_apply(rec, data = avonet_subset, tree = tree_jetz,
                           species_col = "Species1", drop_unresolved = TRUE)
nrow(aligned$data)
ape::Ntip(aligned$tree)
```

## What it does

**Core matching** --- A four-stage cascade resolves names progressively:

1. **Exact** --- identical strings
2. **Normalised** --- case, whitespace, and underscore differences
3. **Synonym** --- taxonomic synonym lookup via local databases (optional)
4. **Fuzzy** --- Levenshtein-based similarity with genus pre-filtering (~100x
   faster than naive pairwise comparison)

**Diagnostics** --- every decision is recorded in a mapping table. Functions
like `reconcile_summary()`, `reconcile_plot()`, `reconcile_report()`, and
`reconcile_suggest()` make it easy to audit matches, visualise coverage, and
find near-misses for unresolved species.

**Multi-tree support** --- `reconcile_to_trees()` matches a dataset against
several trees at once, and `reconcile_diff()` compares results across trees
or parameter settings.

**Crosswalks and overrides** --- `reconcile_crosswalk()` converts published
taxonomy crosswalks (e.g., BirdLife--BirdTree) into override tables.
`reconcile_override()` and `reconcile_override_batch()` let you inject manual
corrections at any stage.

**Advanced** --- `reconcile_augment()` grafts unresolved species onto a tree
using genus-level placement. `reconcile_splits_lumps()` detects taxonomic
splits and lumps from synonym resolution results.

## Typical workflow

```
Load data + tree
       |
  reconcile_tree()
       |
  Review: reconcile_summary(), reconcile_plot(), reconcile_report()
       |
  Fix: reconcile_override(), reconcile_suggest()
       |
  reconcile_apply()
       |
  Aligned data + pruned tree --> PGLS / PGLMM
```

## Vignettes

- [Getting Started](https://itchyshin.github.io/prepR4pcm/docs/articles/getting-started.html)
  ([source](vignettes/getting-started.Rmd))
  --- core concepts and a minimal worked example
- [Bird Trait Workflow](https://itchyshin.github.io/prepR4pcm/docs/articles/bird-workflow.html)
  ([source](vignettes/bird-workflow.Rmd))
  --- a realistic multi-dataset, multi-tree analysis pipeline

## Citation

If you use prepR4pcm in your research, please cite:

```r
citation("prepR4pcm")
```

## License

MIT
