---
title: "From Raw Data to PGLS: A Complete Bird Trait Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Raw Data to PGLS: A Complete Bird Trait Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a real comparative biology workflow using
avian trait data and a phylogenetic tree. The same approach applies to
any taxon group --- mammals, fish, amphibians, plants --- the functions
are fully taxon-agnostic.

## Setup

```{r setup}
library(prepR4pcm)
```

# Part I: Core Workflow

The typical workflow has four steps: load your data and tree, reconcile
the names, produce aligned objects, and run your analysis.

## Step 1: Load data and tree

```{r load-data}
data(avonet_subset)    # AVONET morphological traits (Tobias et al. 2022)
data(tree_jetz)    # Jetz et al. (2012) phylogeny, Corvoidea + allies

cat(sprintf("Data: %d species\n", nrow(avonet_subset)))
cat(sprintf("Tree: %d tips\n", ape::Ntip(tree_jetz)))

# The data uses spaces; the tree uses underscores
head(avonet_subset$Species1, 3)
head(tree_jetz$tip.label, 3)
```

These formatting differences --- spaces vs underscores, minor spelling
variants, taxonomic synonyms --- are exactly what prepR4pcm resolves.

## Step 2: Reconcile data against the tree

```{r reconcile-tree}
result <- reconcile_tree(
  x = avonet_subset,
  tree = tree_jetz,
  x_species = "Species1",
  authority = NULL       # skip synonym lookup for speed
)
print(result)
```

The reconciliation object records every name-matching decision.
Inspect the mapping to see what happened:

```{r mapping}
mapping <- reconcile_mapping(result)

# Match type breakdown
table(mapping$match_type)

# Show normalised matches (formatting differences resolved automatically)
norm <- mapping[mapping$match_type == "normalized",
                c("name_x", "name_y", "notes")]
if (nrow(norm) > 0) head(norm, 5)

# Unresolved: in data but not in tree
unresolved <- mapping[mapping$match_type == "unresolved" & mapping$in_x, ]
cat(sprintf("\nSpecies in data but not in tree: %d\n", nrow(unresolved)))
```

For a detailed report:

```{r summary, eval = FALSE}
reconcile_summary(result, detail = "mismatches_only")
```

## Step 3: Produce aligned objects

Drop unresolved species to get a matched data frame and tree, ready
for comparative analysis:

```{r apply}
aligned <- reconcile_apply(
  result,
  data = avonet_subset,
  tree = tree_jetz,
  species_col = "Species1",
  drop_unresolved = TRUE
)

cat(sprintf("Aligned data: %d rows\nAligned tree: %d tips\n",
            nrow(aligned$data), ape::Ntip(aligned$tree)))
```

## Step 4: Run a comparative analysis

With aligned data and tree, you are ready for any phylogenetic
comparative method. Here are two common approaches.

### Phylogenetic generalised least squares (PGLS)

PGLS accounts for shared evolutionary history when estimating
regression parameters:

```{r pgls, message = FALSE, warning = FALSE}
library(caper)

# reconcile_apply() aligns names so data$Species1 matches tree tip labels
cd <- comparative.data(aligned$tree, aligned$data,
                       names.col = "Species1", vcv = TRUE)

# PGLS: body mass ~ wing length
model_pgls <- pgls(log(Mass) ~ log(Wing.Length), data = cd)
summary(model_pgls)
```

### Phylogenetic generalised linear mixed model (PGLMM)

When you need random effects beyond phylogeny or want a Bayesian
framework, use a PGLMM. The MCMCglmm package fits Bayesian
phylogenetic mixed models:

```{r pglmm, message = FALSE, warning = FALSE, results = "hide"}
library(MCMCglmm)

# Species column as the phylogenetic grouping factor
aligned$data$phylo <- aligned$data$Species1

# Inverse phylogenetic covariance matrix
# Replace any zero-length branches (can arise after pruning)
tree_mcmc <- aligned$tree
tree_mcmc$edge.length[tree_mcmc$edge.length == 0] <- 1e-6
inv_phylo <- inverseA(tree_mcmc, nodes = "ALL", scale = FALSE)

# PGLMM: continuous response
prior <- list(R = list(V = 1, nu = 0.002),
              G = list(G1 = list(V = 1, nu = 0.002)))

model_mcmc <- MCMCglmm(
  log(Mass) ~ log(Wing.Length) + Trophic.Level,
  random = ~phylo,
  family = "gaussian",
  ginverse = list(phylo = inv_phylo$Ainv),
  data = aligned$data,
  prior = prior,
  nitt = 50000, burnin = 10000, thin = 20,
  verbose = FALSE
)
```

```{r pglmm-summary}
summary(model_mcmc)
```

For categorical responses (e.g., migration status with multiple
categories), see Mizuno et al. (2025, *J. Evol. Biol.*
38:1699--1715) for the multinomial PGLMM approach.

See Hadfield (2010, *J. Stat. Softw.* 33:1--22) for MCMCglmm
details, Hadfield & Nakagawa (2010, *J. Evol. Biol.* 23:494--508)
for phylogenetic quantitative genetics, and Mizuno et al. (2025,
*J. Evol. Biol.* 38:1699--1715) for phylogenetic multinomial
mixed models.

That is the complete core workflow: **load, reconcile, apply, analyse.**

---

# Part II: Advanced Topics

## Reconciling two datasets

If you need to harmonise species names across two trait datasets
*before* matching to a tree, use `reconcile_data()`:

```{r reconcile-data}
data(nesttrait_subset)   # Nest traits (Chia et al. 2023)

rec_data <- reconcile_data(
  x = nesttrait_subset,
  y = avonet_subset,
  x_species = "Scientific_name",
  y_species = "Species1",
  authority = NULL,
  quiet = TRUE
)
print(rec_data)
```

Once reconciled, merge the two datasets into a single data frame:

```{r merge-data}
merged <- reconcile_merge(
  rec_data,
  data_x = nesttrait_subset,
  data_y = avonet_subset,
  species_col_x = "Scientific_name",
  species_col_y = "Species1"
)
cat(sprintf("Merged: %d rows, %d columns\n", nrow(merged), ncol(merged)))
```

## Using a taxonomy crosswalk

When your data and tree use different taxonomies (e.g., BirdLife data
against a BirdTree phylogeny), a curated crosswalk can resolve names
that automated synonym resolution misses.

A crosswalk is simply a table mapping names from one system to another.
prepR4pcm includes the BirdLife-BirdTree crosswalk as an example:

```{r crosswalk}
data(crosswalk_birdlife_birdtree)
table(crosswalk_birdlife_birdtree$Match.type)
```

Convert it to an overrides table and pass it to `reconcile_tree()`:

```{r make-overrides}
overrides <- reconcile_crosswalk(
  crosswalk_birdlife_birdtree,
  from_col = "Species1",
  to_col = "Species3",
  match_type_col = "Match.type"
)

# Re-reconcile with overrides
result_xw <- reconcile_tree(
  x = avonet_subset,
  tree = tree_jetz,
  x_species = "Species1",
  authority = NULL,
  overrides = overrides
)

# Compare: how many more matches with the crosswalk?
cat(sprintf("Without crosswalk: %d matched\n",
            sum(result$mapping$in_x & result$mapping$in_y, na.rm = TRUE)))
cat(sprintf("With crosswalk:    %d matched\n",
            sum(result_xw$mapping$in_x & result_xw$mapping$in_y, na.rm = TRUE)))
```

**When do you need a crosswalk?** Only when your data and tree follow
different naming authorities *and* a curated mapping exists. For most
use cases, the automatic cascade (exact → normalised → synonym) is
sufficient.

You can also build your own overrides manually --- it is just a data
frame with `name_x`, `name_y`, and optionally `user_note` columns:

```{r manual-overrides, eval = FALSE}
my_overrides <- data.frame(
  name_x    = c("Old name A", "Old name B"),
  name_y    = c("Tree name A", "Tree name B"),
  user_note = c("Reclassified in 2023", "Spelling correction")
)
result <- reconcile_tree(my_data, my_tree, overrides = my_overrides)
```

## Reconciling against multiple trees

For sensitivity analyses across phylogenies, `reconcile_to_trees()`
reconciles one dataset against several trees in one call:

```{r multi-tree}
data(tree_clements25)  # Clements 2025 tree

results <- reconcile_to_trees(
  x = avonet_subset,
  trees = list(
    jetz      = tree_jetz,
    clements  = tree_clements25
  ),
  x_species = "Species1",
  authority = NULL
)

# Compare overlap across trees
sapply(results, function(r) {
  c(matched = sum(r$mapping$in_x & r$mapping$in_y, na.rm = TRUE),
    unresolved_x = r$counts$n_unresolved_x)
})
```

## Fuzzy matching for typos

Enable fuzzy matching to catch likely typos in species names:

```{r fuzzy, eval = FALSE}
result <- reconcile_tree(
  x = my_data,
  tree = my_tree,
  fuzzy = TRUE,              # enable fuzzy matching
  fuzzy_threshold = 0.9,     # minimum similarity (0-1)
  resolve = "flag"           # flag low-confidence matches for review
)

# Check flagged matches
flagged <- reconcile_mapping(result)
flagged[flagged$match_type == "flagged", c("name_x", "name_y", "match_score")]
```

## Tree augmentation for missing species

When the tree has fewer species than the data, `reconcile_apply()`
drops the unresolved species. This loses statistical power and can
bias the sample. `reconcile_augment()` grafts the missing species
onto the tree using genus-level placement:

```{r augment}
aug <- reconcile_augment(
  result,
  tree_jetz,
  where = "genus",                # sister to a random congener

  branch_length = "congener_median",  # median terminal branch of congeners
  seed = 42,                      # for reproducibility
  quiet = TRUE
)

cat(sprintf("Original tips: %d\nAugmented tips: %d\n",
            ape::Ntip(aug$original), ape::Ntip(aug$tree)))
cat(sprintf("Added: %d | Skipped (no congener): %d\n",
            nrow(aug$augmented), nrow(aug$skipped)))

# Which species were added, and where?
if (nrow(aug$augmented) > 0) head(aug$augmented[, c("species", "placed_near", "branch_length")])
```

Use the augmented tree in downstream analyses:

```{r augment-apply, eval = FALSE}
aligned_aug <- reconcile_apply(
  result,
  data = avonet_subset,
  tree = aug$tree,
  species_col = "Species1",
  drop_unresolved = TRUE
)
```

**Important caveat.** Genus-level placement assumes the missing
species diverged similarly to its congeners, which may not hold.
Always report which species were augmented and run sensitivity
analyses comparing results with and without them.

## Exporting to files

Write aligned data, tree, and the full mapping table to disk:

```{r export, eval = FALSE}
reconcile_export(
  result,
  data = avonet_subset,
  tree = tree_jetz,
  species_col = "Species1",
  dir = "output",
  prefix = "avonet_jetz"
)
# Writes: avonet_jetz_data.csv, avonet_jetz_tree.nex, avonet_jetz_mapping.csv
```

## HTML reports

Generate a self-contained HTML report documenting every name-matching
decision. Useful for sharing with collaborators or archiving alongside
your analysis:

```{r report, eval = FALSE}
reconcile_report(result, file = "reconciliation_report.html")
```

The report includes match counts, detail tables for each match type
(normalised, synonym, fuzzy, flagged), and lists of unresolved species.

## Key points

1. **Taxon-agnostic.** This workflow works for any group --- mammals,
   fish, amphibians, plants --- as long as you have a data frame and a
   phylogenetic tree.

2. **Provenance.** Every name-matching decision is recorded in the
   `reconciliation` object. Use `reconcile_summary()` for a
   human-readable report or `reconcile_mapping()` for the full table.

3. **Crosswalks are optional.** Most users do not need them. The
   automatic cascade handles formatting differences and synonyms.
   Crosswalks help when two well-known naming authorities disagree.

4. **Tree augmentation.** When the tree is incomplete,
   `reconcile_augment()` grafts missing species using congener
   placement --- but always run sensitivity analyses with and without
   augmented tips.

5. **Sensitivity.** `reconcile_to_trees()` makes it easy to run the
   same analysis across multiple phylogenies.

6. **Merging.** `reconcile_merge()` joins two reconciled datasets into
   a single analysis-ready data frame, using the mapping as the join key.

7. **Reports.** `reconcile_report()` generates a self-contained HTML
   report suitable for sharing or archiving.

8. **Visualisation.** `reconcile_plot()` produces a bar or pie chart of
   match composition. `reconcile_suggest()` shows the closest fuzzy
   candidates for unresolved species.

9. **Comparison.** `reconcile_diff()` compares two reconciliation runs
   side by side --- e.g., before and after adding a crosswalk.

## Data sources

- **AVONET**: Tobias et al. (2022) *Ecology Letters* 25:581--597.
  doi:10.1111/ele.13898
- **NestTrait v2**: Chia et al. (2023) *Scientific Data* 10:444.
  doi:10.1038/s41597-023-02360-9
- **Plumage lightness**: Delhey et al. (2019) *American Naturalist*
  194:13--27. doi:10.1086/703588
- **Jetz tree**: Jetz et al. (2012) *Nature* 491:444--448.
  doi:10.1038/nature11631
- **Clements 2025**: Clements et al. (2025) eBird/Clements Checklist.
- **BirdLife-BirdTree crosswalk**: Tobias et al. (2022).

## References

- Hadfield, J.D. (2010) MCMC methods for multi-response generalized
  linear mixed models: the MCMCglmm R package. *Journal of Statistical
  Software* 33:1--22. doi:10.18637/jss.v033.i02
- Hadfield, J.D. & Nakagawa, S. (2010) General quantitative genetic
  methods for comparative biology: phylogenies, taxonomies and
  multi-trait models for continuous and categorical characters.
  *Journal of Evolutionary Biology* 23:494--508.
  doi:10.1111/j.1420-9101.2009.01915.x
- Mizuno, A., Drobniak, S.M., Williams, C., Lagisz, M. & Nakagawa, S.
  (2025) Promoting the use of phylogenetic multinomial generalised
  mixed-effects model to understand the evolution of discrete traits.
  *Journal of Evolutionary Biology* 38:1699--1715.
  doi:10.1093/jeb/voaf116
- Norman, K.E., Chamberlain, S. & Boettiger, C. (2020) taxadb: A
  high-performance local taxonomic database interface. *Methods in
  Ecology and Evolution* 11:1153--1159. doi:10.1111/2041-210X.13440
- Paradis, E. & Schliep, K. (2019) ape 5.0: an environment for modern
  phylogenetics and evolutionary analyses in R. *Bioinformatics*
  35:526--528. doi:10.1093/bioinformatics/bty633
