---
title: "From Raw Data to PGLS: A Complete Bird Trait Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Raw Data to PGLS: A Complete Bird Trait Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a real comparative biology workflow using
avian trait data and phylogenetic trees. The same approach applies to
any taxon group --- mammals, fish, amphibians, plants --- the functions
are fully taxon-agnostic.

## Setup

```{r setup}
library(prepR4pcm)
```

## Step 1: Load the data

prepR4pcm ships with subsets of three published bird trait datasets and
two phylogenetic trees, selected to illustrate common name-matching
challenges.

```{r load-data}
data(avonet_subset)        # AVONET morphological traits (Tobias et al. 2022)
data(nesttrait_subset)     # Nest traits (Chia et al. 2023)
data(delhey_subset)        # Plumage lightness (Delhey et al. 2019)
data(tree_jetz_200)        # Jetz et al. (2012) tree, pruned
data(tree_clements25_200)  # Clements 2025 tree, pruned

str(avonet_subset[, 1:5])
str(nesttrait_subset[, 1:5])
head(delhey_subset$TipLabel)  # Note: underscores, not spaces
```

Notice the naming differences:

- AVONET uses `Species1` with spaces: `"Abeillia abeillei"`
- NestTrait uses `Scientific_name` with spaces: `"Abeillia abeillei"`
- Delhey uses `TipLabel` with underscores: `"Abeillia_abeillei"`
- Tree tip labels use underscores: `"Abeillia_abeillei"`

These formatting differences are exactly what prepR4pcm resolves
automatically.

## Step 2: Reconcile datasets against each other

Before matching to a tree, harmonise species names across trait
datasets. Here we reconcile NestTrait names against AVONET names:

```{r reconcile-data}
rec_nest <- reconcile_data(
  x = nesttrait_subset,
  y = avonet_subset,
  x_species = "Scientific_name",
  y_species = "Species1",
  authority = NULL,          # skip synonym lookup for speed
  quiet = FALSE
)
print(rec_nest)
```

Most names match exactly because both datasets use BirdLife taxonomy.
The few mismatches are typically formatting differences caught by the
normalisation stage.

## Step 3: Use a taxonomy crosswalk

When matching data to a phylogenetic tree built from a different
taxonomy (e.g., BirdLife data against a BirdTree phylogeny), you can
load a curated crosswalk as overrides:

```{r crosswalk}
data(crosswalk_birdlife_birdtree)
head(crosswalk_birdlife_birdtree)
table(crosswalk_birdlife_birdtree$Match.type)
```

Convert the crosswalk to an overrides table:

```{r make-overrides}
overrides <- reconcile_crosswalk(
  crosswalk_birdlife_birdtree,
  from_col = "Species1",
  to_col = "Species3",
  match_type_col = "Match.type"
)
head(overrides)
```

This tells prepR4pcm: "when you see `Species1` in the data, match it
to `Species3` in the tree." The crosswalk handles splits, lumps, and
renames that automated synonym resolution might miss.

## Step 4: Reconcile data against one tree

```{r reconcile-tree}
rec_jetz <- reconcile_tree(
  x = avonet_subset,
  tree = tree_jetz_200,
  x_species = "Species1",
  authority = NULL,
  overrides = overrides,
  quiet = FALSE
)
print(rec_jetz)
```

### Inspect the mapping

```{r mapping}
mapping <- reconcile_mapping(rec_jetz)

# Show a few normalised matches
mapping[mapping$match_type == "normalized", c("name_x", "name_y", "notes")]
```

### Detailed report

```{r summary, eval = FALSE}
reconcile_summary(rec_jetz, detail = "mismatches_only")
```

## Step 5: Reconcile against multiple trees

The typical PCM workflow requires separate tree-compatible datasets
for each available phylogeny. `reconcile_to_trees()` handles this in
one call:

```{r multi-tree}
results <- reconcile_to_trees(
  x = avonet_subset,
  trees = list(
    jetz      = tree_jetz_200,
    clements  = tree_clements25_200
  ),
  x_species = "Species1",
  authority = NULL,
  overrides = overrides,
  quiet = FALSE
)

# Compare overlap across trees
sapply(results, function(r) {
  c(matched = sum(r$mapping$in_x & r$mapping$in_y, na.rm = TRUE),
    unresolved_x = r$counts$n_unresolved_x)
})
```

## Step 6: Produce aligned objects

```{r apply}
aligned <- reconcile_apply(
  results$jetz,
  data = avonet_subset,
  tree = tree_jetz_200,
  species_col = "Species1",
  drop_unresolved = TRUE
)

cat(sprintf("Data: %d rows\nTree: %d tips\n",
            nrow(aligned$data), ape::Ntip(aligned$tree)))
```

## Step 7: Export to files

```{r export, eval = FALSE}
reconcile_export(
  results$jetz,
  data = avonet_subset,
  tree = tree_jetz_200,
  species_col = "Species1",
  dir = "output",
  prefix = "avonet_jetz",
  tree_format = "nexus"
)
# Writes:
#   output/avonet_jetz_data.csv
#   output/avonet_jetz_tree.nex
#   output/avonet_jetz_mapping.csv
```

## Step 8: Run a comparative analysis

With aligned data and tree, you are ready for any phylogenetic
comparative method:

```{r pgls, eval = FALSE}
library(caper)

# Prepare for caper
aligned$data$species <- gsub(" ", "_", aligned$data$Species1)
cd <- comparative.data(aligned$tree, aligned$data,
                       names.col = "species", vcv = TRUE)

# PGLS: body mass ~ wing length
model <- pgls(log(Mass) ~ log(Wing.Length), data = cd)
summary(model)
```

## Key points

1. **Taxon-agnostic.** This workflow works for any group --- mammals,
   fish, amphibians, plants --- as long as you have a data frame and a
   phylogenetic tree.

2. **Provenance.** Every name-matching decision is recorded in the
   `reconciliation` object. Use `reconcile_summary()` for a
   human-readable report or `reconcile_mapping()` for the full table.

3. **Crosswalks.** When you have a curated taxonomy mapping (BirdLife
   to BirdTree, WFO to GBIF, etc.), use `reconcile_crosswalk()` to
   convert it to overrides. This is often more reliable than automated
   synonym resolution for well-studied groups.

4. **Multiple trees.** `reconcile_to_trees()` reconciles one dataset
   against several trees in one call, making sensitivity analyses
   across phylogenies straightforward.

## Data sources

- **AVONET**: Tobias et al. (2022) *Ecology Letters* 25:581--597.
  doi:10.1111/ele.13898
- **NestTrait v2**: Chia et al. (2023) *Scientific Data* 10:444.
  doi:10.1038/s41597-023-02360-9
- **Plumage lightness**: Delhey et al. (2019) *American Naturalist*
  194:13--27. doi:10.1086/703588
- **Jetz tree**: Jetz et al. (2012) *Nature* 491:444--448.
  doi:10.1038/nature11631
- **Clements 2025**: Clements et al. (2025) eBird/Clements Checklist.
- **BirdLife-BirdTree crosswalk**: Tobias et al. (2022).
