---
title: "Comparing tree backends — when do they agree?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing tree backends — when do they agree?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`prepR4pcm` ships five tree-retrieval backends. They draw on
different reference trees, different taxonomies, and different
calibration approaches. So they disagree. The question is *how
much*, and *where*.

This vignette walks through `pr_tree_compare()` and shows how to
use it to make a defensible backend choice for your dataset — or to
report which backend choice you made and what the alternatives
would have given.

## Setup

```{r setup, eval = TRUE}
library(prepR4pcm)
# install.packages(c("rotl", "fishtree"))
# pak::pak(c("eliotmiller/clootl", "daijiang/rtrees", "phylotastic/datelife"))
```

Check what's installed and reachable in your session:

```{r status, eval = TRUE}
pr_get_tree_status()
```

## The recipe

Pick a small species list. Retrieve the same set from two or three
backends. Compare.

```{r retrieve}
species <- c("Salmo salar", "Esox lucius", "Oncorhynchus mykiss",
             "Gadus morhua", "Thunnus thynnus")

# Three backends that all cover fish
res_rotl     <- pr_get_tree(species, source = "rotl")
res_fishtree <- pr_get_tree(species, source = "fishtree")
res_rtrees   <- pr_get_tree(species, source = "rtrees", taxon = "fish")
```

The three results are independent. Now compare:

```{r compare}
cmp <- pr_tree_compare(
  rotl     = res_rotl,
  fishtree = res_fishtree,
  rtrees   = res_rtrees
)
cmp
```

The print method shows three pairwise matrices. Reading them:

- **Tip-set Jaccard** — fraction of species both trees place. 1.0 =
  same species set. 0.0 = no overlap. If any pairwise value is < 1,
  you're not comparing apples to apples.
- **Robinson-Foulds distance** — number of internal edges that
  differ between the two trees, restricted to their shared tip set.
  Lower = trees agree more.
- **Branch-length correlation** — Pearson correlation of sorted
  edge lengths after pruning to the shared subtree. Useful as a
  rough check on whether two chronograms agree on rate scale.

## What to do with the result

### When all three agree (Jaccard ≈ 1, RF small)

Great. Use whichever backend best matches your taxonomic / temporal
needs. Document the choice. Call `pr_cite_tree()` to capture the
citation:

```{r cite}
pr_cite_tree(res_fishtree, format = "markdown")
```

### When tip sets differ (Jaccard < 1)

The backends disagree on which species are valid. Inspect the
unique-to lists:

```{r unique-to}
cmp$unique_to$rotl     # species rotl placed but the others didn't
cmp$unique_to$fishtree # species fishtree placed but the others didn't
cmp$shared_tips        # species placed by all backends
```

Common causes:
- **Taxonomy difference.** rotl uses Open Tree taxonomy; fishtree
  uses its own; clootl uses Clements. The *same* species may have
  different canonical names.
- **Coverage gap.** A species is in one reference tree but not
  another (often a recently described species).
- **Synonymy.** One backend matches via synonymy, another doesn't.

Mitigation: re-run with `tnrs = "always"` to harmonise names via
Open Tree's TNRS first. See `?pr_get_tree`.

### When topologies disagree (RF large)

The trees agree on which species exist but disagree on how they're
related. This is the genuinely interesting case.

- For exploratory PCMs, run your model on each tree and report the
  range of estimates.
- For a definitive analysis, pick the backend with the strongest
  underlying paper for your taxon (fishtree for fish, clootl for
  birds, datelife for everything else when you need calibration).

The companion package
[pigauto](https://itchyshin.github.io/pigauto/) is built for this:
hand it the multiPhylo of every backend's tree, fit your model on
each, and pool the results via Rubin's rules. The resulting
intervals reflect the tree-choice uncertainty honestly.

```{r pigauto-multi-backend}
# Fitting across backends (not just within a backend's posterior)
all_trees <- structure(list(res_rotl$tree, res_fishtree$tree,
                            res_rtrees$tree[[1]]),
                       class = "multiPhylo")
# pigauto::multi_impute_trees(rec$aligned, all_trees, m_per_tree = 5)
```

### When branch-length correlation is low

The trees agree on topology but not on rate scale. Usually one is
time-calibrated and the other isn't:

- `rotl` returns the synthesis tree with no calibration.
- `rtrees` grafts onto a reference tree; calibration is whatever the
  reference uses.
- `fishtree` and `clootl` and `datelife` all return time-calibrated
  trees.

If your downstream model expects calibrated branches (most BM /
OU / lambda models do), use a calibrated backend. If the only
calibrated option for your taxon is `datelife`, see
`?pr_date_tree` for dating an existing topology.

## Caching the comparisons

Each retrieval is slow. If you're going to compare backends, set
`cache = TRUE` so re-running the vignette is instant:

```{r cache}
res_rotl     <- pr_get_tree(species, source = "rotl",
                             cache = TRUE)
res_fishtree <- pr_get_tree(species, source = "fishtree",
                             cache = TRUE)
res_rtrees   <- pr_get_tree(species, source = "rtrees",
                             taxon  = "fish", cache = TRUE)

# See what's cached
pr_tree_cache_status()

# Wipe just the rotl entries (e.g. after an OTT version refresh)
pr_tree_cache_clear(source = "rotl", confirm = FALSE)
```

By default the cache lives at `tools::R_user_dir()`. Pass a
project-local directory if you want the cache committed alongside
the analysis:

```{r project-cache}
pr_tree_cache_dir("./.tree-cache")
```

## What `pr_tree_compare()` doesn't do (yet)

- **Bipartition-matched branch-length correlation.** The current
  implementation correlates *sorted* edge lengths, which is a coarse
  but defensible "do they agree on rate scale" check. A proper
  bipartition-matched correlation is a future-round refinement.
- **Visualisation.** Pair with `phytools::cophyloplot()` or
  `phangorn::densiTree()` for visual comparison of two or many
  trees.
- **Three-or-more-tree consensus.** For "what's the median tree
  across these three backends?", use
  `phangorn::consensus(c(res_rotl$tree, res_fishtree$tree, res_rtrees$tree[[1]]))`.

## See also

- `?pr_tree_compare` — full reference for the comparison function.
- `?pr_get_tree` — retrieval entry point.
- `?pr_get_tree_status` — backend health probe.
- `?pr_tree_cache_dir` — cache management.
- The
  [posterior-tree pipeline vignette](posterior-tree-pipeline.html)
  for the prepR4pcm → pigauto integration.
