---
title: "Comparing tree backends — when do they agree?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing tree backends — when do they agree?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

`prepR4pcm` ships five tree-retrieval backends. They draw on
different reference trees, different taxonomies, and different
calibration approaches. So they disagree. The question is *how
much*, and *where*.

This vignette walks through `pr_tree_compare()` and shows how to
use it to make a defensible backend choice for your dataset — or to
report which backend choice you made and what the alternatives
would have given.

## Setup

Comparison requires multiple tree-retrieval backends installed. Each
backend lives in a different package because each draws on a
different reference tree or taxonomy:

- **`rotl`** (CRAN): client for the Open Tree of Life synthesis tree
  — the only backend with universal taxon coverage. Required for
  `source = "rotl"`.
- **`fishtree`** (CRAN): wraps the Rabosky et al. (2018)
  time-calibrated ray-finned fish phylogeny. Required for
  `source = "fishtree"`.
- **`rtrees`** (GitHub only): grafts species onto taxon-specific
  mega-trees (mammals, birds, fish, amphibians, etc.). Required for
  `source = "rtrees"`.
- **`clootl`** (GitHub only): supplies the Clements bird-taxonomy
  phylogeny. Required for `source = "clootl"`.
- **`datelife`** (GitHub only): synthesis chronograms from many
  per-clade published trees. Required for `source = "datelife"`.

Install whichever you need. CRAN packages first, then GitHub-only
backends via `pak`:

```{r setup, eval = TRUE}
library(prepR4pcm)
# install.packages(c("rotl", "fishtree"))                 # CRAN
# pak::pak("daijiang/rtrees")                             # GitHub
# pak::pak("eliotmiller/clootl")                          # GitHub
# pak::pak("phylotastic/datelife")                        # GitHub (heavy)
```

Check what's installed and reachable in your session:

```{r status, eval = TRUE}
pr_get_tree_status()
```

## The recipe

Pick a small species list. Retrieve the same set from two or three
backends. Compare.

```{r retrieve}
species <- c("Salmo salar", "Esox lucius", "Oncorhynchus mykiss",
             "Gadus morhua", "Thunnus thynnus")

# Three backends that all cover fish
res_rotl     <- pr_get_tree(species, source = "rotl")
res_fishtree <- pr_get_tree(species, source = "fishtree")
res_rtrees   <- pr_get_tree(species, source = "rtrees", taxon = "fish")
```

**About the `TNRS replaced ...` warning.** `pr_get_tree()` runs the
Open Tree of Life Taxonomic Name Resolution Service (TNRS) on the
input names before passing them to the backend, so that minor
spelling or capitalisation differences don't cause a name to fall
out of the tree silently. When TNRS finds a canonical form that
differs from the input (e.g. an old binomial replaced by the
current synonym), the wrapper substitutes the canonical form and
emits a warning listing each replacement, so the substitution is
visible in the run log. The retrieval still succeeds; the warning is
informational.

The three results are independent. Now compare:

```{r compare}
cmp <- pr_tree_compare(
  rotl     = res_rotl,
  fishtree = res_fishtree,
  rtrees   = res_rtrees
)
cmp
```

The print method shows up to three pairwise matrices, depending on
what is computable for your set of trees. Reading them:

- **Tip-set Jaccard** — fraction of species both trees place. 1.0 =
  same species set. 0.0 = no overlap. If any pairwise value is < 1,
  the two trees do not cover the same species set, so the
  topological comparison below is restricted to the species both
  trees include.
- **Robinson-Foulds distance** — number of internal edges that
  differ between the two trees, restricted to their shared tip set.
  Lower = trees agree more. **`NA` off-diagonal** means the
  shared-tip subtree has fewer than four taxa (Robinson-Foulds is
  undefined below four shared tips), or one of the input trees
  could not be pruned to the shared subtree without becoming
  degenerate. Diagonal values are always 0 (a tree against itself).
- **Branch-length correlation** — Pearson correlation of sorted
  edge lengths after pruning to the shared subtree. Useful as a
  rough check on whether two chronograms use the same
  branch-length scale (e.g. millions of years vs unit-length
  placeholders). **This matrix is only printed when at least one
  pair has computable branch lengths on both sides.** If `rotl`
  trees (which carry unit-length placeholder branches) are part of
  the comparison, the matrix may collapse to NA and the print
  method drops it; that is why you sometimes see only two matrices
  rather than three.

## What actually works (status table)

Backends differ in coverage and in what `n_tree > 1` does. Verified
on a clean macOS R 4.4 install on 2026-05-01. Column meanings:

- **Backend** — the value you pass as `source = "..."` to
  `pr_get_tree()`.
- **Single tree** — does `pr_get_tree(..., n_tree = 1)` (the
  default) return a single `ape::phylo` object cleanly?
- **Multi-tree (`n_tree > 1`)** — does the backend support
  returning a posterior or sample of trees? Some backends ship a
  single mega-tree and ignore `n_tree`; others support a true
  posterior sample.
- **Notes** — backend-specific gotchas worth knowing before you
  trust the result.

| Backend | Single tree | Multi-tree (`n_tree > 1`) | Notes |
|---|---|---|---|
| `rotl` | ✅ | n/a — synthesis is a single tree | Returns lowercase tip labels (e.g. `salmo salar`) because Open Tree taxonomy names are case-folded; reconcile against the tree if your input mixes cases. |
| `rtrees` | ✅ | ✅ for taxa whose mega-tree is a posterior (e.g. `taxon = "bird"`, `"mammal"` → 100 trees); returns 1 tree for taxa with a single mega-tree (e.g. `taxon = "fish"`). | `n_tree` is **informational only** — `rtrees::get_tree()` has no `n_tree` argument, so the count is fixed by the chosen mega-tree. |
| `clootl` | ✅ | ❌ unless the AvesData repo is installed (run `clootl::get_avesdata_repo(".")` once). `n_tree = 1` calls `clootl::extractTree()` and works out of the box; `n_tree > 1` calls `clootl::sampleTrees(count = …)` and needs AvesData. | The single-tree path uses the v1.6 / 2025 taxonomy bundled with `clootl`. Posterior sampling caps at 100 upstream. |
| `fishtree` | ✅ | ✅ — `n_tree > 1` switches to `fishtree::fishtree_complete_phylogeny()` and returns the requested count. | Time-calibrated. |
| `datelife` | likely ✅ | likely ✅ | Untested in this run because `datelife` is in `Enhances` (heavy Bioconductor / BOLD deps). Install with `pak::pak("phylotastic/datelife")`. |
| `pr_date_tree()` | likely ✅ | likely ✅ | Same dependency story as `datelife`. |

If you hit a **broken** row above on a fresh install, please open an
issue at
[itchyshin/prepR4pcm](https://github.com/itchyshin/prepR4pcm/issues)
with your `pr_get_tree_status()` output and the error.

## Where are the VertLife trees?

A common question: *"I want the VertLife / Upham et al. 2019 mammal
posterior — which backend has it?"*

Short answer: **`source = "rtrees"`** has them. `rtrees` depends on
the [`megatrees`](https://github.com/daijiang/megatrees) package,
which ships 100 randomly-sampled posterior trees from each of the
major VertLife datasets (mammals: Upham et al. 2019; amphibians:
Jetz & Pyron 2018; squamates: Tonini et al. 2016; sharks: Stein et
al. 2018; birds: Jetz et al. 2012). When you call
`pr_get_tree(species, source = "rtrees", taxon = "mammal")`, the
returned `multiPhylo` is exactly that 100-tree subset of the
VertLife posterior, grafted to your species set.

If you need the **full 10,000-tree posterior** (rather than the
100-tree subset), you currently have to download the source archive
manually from [vertlife.org](https://vertlife.org/data/) (each
archive is 0.5–2 GB). A future round may add a `source = "vertlife"`
backend that automates this caching step; for now, the 100-tree
subset via `rtrees` is what you get out of the box, which is
sufficient for the great majority of phylogenetic comparative
analyses.

## Note on tip labels before comparing

The three backends use different conventions for tip labels:

- **`rotl`** appends an Open Tree OTT id to each tip
  (e.g. `Salmo_salar_ott854188`) and lower-cases the binomial.
- **`fishtree`** returns the canonical binomial with spaces
  (`Salmo salar`).
- **`rtrees`** returns the canonical binomial with underscores
  (`Salmo_salar`).

`pr_tree_compare()` matches tips by **case- and separator-folded
binomial** (it strips `_ott<digits>`, converts underscores to
spaces, and lower-cases) before computing Jaccard, so you do not
need to clean the labels yourself for the comparison to work. The
folded form is used only for matching; the original tip labels are
preserved in each tree.

If you intend to use one of the returned trees downstream (e.g.
feed it to `pr_phylo_cor()` and then to `metafor::rma.mv()`), you
will usually want to strip those suffixes yourself with
`gsub("_ott\\d+", "", tree$tip.label)` first. See the
*meta-analysis-with-rotl* vignette for that workflow.

## What to do with the result

### When the backends return the same species set and similar topology (Jaccard ≈ 1, RF small)

The backends return the same species set (Jaccard ≈ 1) and broadly
similar topologies (low Robinson-Foulds distance). Pick whichever
backend best matches your taxonomic / temporal needs. Document the
choice. Call `pr_cite_tree()` to capture the citation:

```{r cite}
pr_cite_tree(res_fishtree, format = "markdown")
```

### When tip sets differ (Jaccard < 1)

The backends disagree on which species are valid. Inspect the
unique-to lists:

```{r unique-to}
cmp$unique_to$rotl     # species rotl placed but the others didn't
cmp$unique_to$fishtree # species fishtree placed but the others didn't
cmp$shared_tips        # species placed by all backends
```

Common causes:
- **Taxonomy difference.** rotl uses Open Tree taxonomy; fishtree
  uses its own; clootl uses Clements. The *same* species may have
  different canonical names.
- **Coverage gap.** A species is in one reference tree but not
  another (often a recently described species).
- **Synonymy.** One backend matches via synonymy, another doesn't.

Mitigation: re-run with `tnrs = "always"` to harmonise names via
Open Tree's TNRS first. See `?pr_get_tree`.

### When topologies disagree (RF large)

The trees agree on which species exist but disagree on how they're
related. This is the genuinely interesting case.

- For exploratory PCMs, run your model on each tree and report the
  range of estimates.
- For a definitive analysis, pick the backend with the strongest
  underlying paper for your taxon (fishtree for fish, clootl for
  birds, datelife for everything else when you need calibration).

The companion package
[pigauto](https://itchyshin.github.io/pigauto/) is built for this:
hand it the multiPhylo of every backend's tree, fit your model on
each, and pool the results via Rubin's rules. The resulting
intervals reflect the tree-choice uncertainty honestly.

```{r pigauto-multi-backend}
# Fitting across backends (not just within a backend's posterior)
all_trees <- structure(list(res_rotl$tree, res_fishtree$tree,
                            res_rtrees$tree[[1]]),
                       class = "multiPhylo")
# pigauto::multi_impute_trees(rec$aligned, all_trees, m_per_tree = 5)
```

### When branch-length correlation is low

The trees agree on topology but not on **branch-length scale**.
Usually that is because one tree is time-calibrated (branch lengths
in millions of years) and the other is not (branch lengths in
unit-length placeholders, or in Grafen-units, or in some other
non-time scale):

- `rotl` returns the synthesis tree with no calibration.
- `rtrees` grafts onto a reference tree; calibration is whatever the
  reference uses.
- `fishtree` and `clootl` and `datelife` all return time-calibrated
  trees.

If your downstream model expects calibrated branches (most BM /
OU / lambda models do), use a calibrated backend. If the only
calibrated option for your taxon is `datelife`, see
`?pr_date_tree` for dating an existing topology.

## Caching the comparisons

Each retrieval is slow. If you're going to compare backends, set
`cache = TRUE` so re-running the vignette is instant:

```{r cache}
res_rotl     <- pr_get_tree(species, source = "rotl",
                             cache = TRUE)
res_fishtree <- pr_get_tree(species, source = "fishtree",
                             cache = TRUE)
res_rtrees   <- pr_get_tree(species, source = "rtrees",
                             taxon  = "fish", cache = TRUE)

# See what's cached
pr_tree_cache_status()

# Wipe just the rotl entries (e.g. after an OTT version refresh)
pr_tree_cache_clear(source = "rotl", confirm = FALSE)
```

By default the cache lives at `tools::R_user_dir()`. Pass a
project-local directory if you want the cache committed alongside
the analysis:

```{r project-cache}
pr_tree_cache_dir("./.tree-cache")
```

## What `pr_tree_compare()` doesn't do (yet)

- **Bipartition-matched branch-length correlation.** The current
  implementation correlates *sorted* edge lengths, which is a coarse
  but defensible "do they agree on rate scale" check. A proper
  bipartition-matched correlation is a future-round refinement.
- **Visualisation.** Pair with `phytools::cophyloplot()` or
  `phangorn::densiTree()` for visual comparison of two or many
  trees.
- **Three-or-more-tree consensus.** For "what's the median tree
  across these three backends?", use
  `phangorn::consensus(c(res_rotl$tree, res_fishtree$tree, res_rtrees$tree[[1]]))`.

## See also

- `?pr_tree_compare` — full reference for the comparison function.
- `?pr_get_tree` — retrieval entry point.
- `?pr_get_tree_status` — backend health probe.
- `?pr_tree_cache_dir` — cache management.
- The
  [posterior-tree pipeline vignette](posterior-tree-pipeline.html)
  for the prepR4pcm → pigauto integration.
