---
title: "Getting Started with prepR4pcm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with prepR4pcm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Before running any phylogenetic comparative analysis — PGLS, phylogenetic
mixed models, ancestral state reconstruction — species names in your data
must match the tip labels in your tree. In practice, they rarely do.
**prepR4pcm** automates this reconciliation, documents every decision, and
produces aligned objects ready for analysis.

## The problem

Mismatches between data and tree arise from:
- **Formatting**: `Homo_sapiens` vs `Homo sapiens`, trailing whitespace,
  attached authority strings
- **Synonymy**: Taxonomic revisions move species between genera
- **Missing names**: Species in the data but not the tree (or vice versa)

Fixing these by hand is tedious, error-prone, and poorly documented.
prepR4pcm solves this with a structured matching cascade: exact match →
normalised match → synonym resolution. Every decision is recorded.

## Installation

```{r install, eval = FALSE}
# Install pak if you don't have it
# install.packages("pak")

# Install prepR4pcm from GitHub
pak::pak("itchyshin/prepR4pcm")
```

```{r setup}
library(prepR4pcm)
```

## Example 1: Reconcile a dataset against a tree

Suppose you have trait data and a phylogenetic tree with slightly different
naming conventions.

```{r example-data}
# Simulated trait data
trait_data <- data.frame(
  species = c(
    "Homo sapiens",
    "Pan_troglodytes",       # underscore instead of space
    "Gorilla gorilla",
    "Pongo pygmaeus",
    "Macaca mulatta",
    "Cebus capucinus"
  ),
  body_mass = c(70, 50, 160, 80, 8, 3),
  brain_mass = c(1.35, 0.39, 0.50, 0.37, 0.11, 0.07)
)

# Simulated phylogenetic tree (built manually for this example)
tree <- ape::read.tree(text = paste0(
  "((((Homo_sapiens:5,Pan_troglodytes:5):3,",
  "Gorilla_gorilla:8):4,Pongo_pygmaeus:12):6,",
  "(Macaca_mulatta:10,Papio_anubis:10):8);"
))
```

Notice the mismatches:
- `Pan_troglodytes` in the data has an underscore; the tree uses underscores too,
  but the data column uses a mix
- `Cebus capucinus` is in the data but not the tree
- `Papio anubis` is in the tree but not the data

```{r reconcile-tree}
result <- reconcile_tree(
  x = trait_data,
  tree = tree,
  x_species = "species",
  authority = NULL,        # skip synonym lookup for this example
  quiet = FALSE
)
```

### Inspect the result

```{r print-result}
print(result)
```

The `reconciliation` object tells you immediately how many names matched
and how. Use `reconcile_mapping()` to see the full table:

```{r mapping}
reconcile_mapping(result)
```

For a detailed report:

```{r summary, eval = FALSE}
reconcile_summary(result)
```

### Apply manual overrides

Suppose you know that `Cebus capucinus` should not be in the analysis.
You can document this decision:
```{r override}
result <- reconcile_override(
  result,
  name_x = "Cebus capucinus",
  name_y = NA,
  action = "reject",
  note = "Not in target phylogeny; exclude from analysis"
)
```

`reconcile_override()` updates the existing reconciliation object in
place — no need to re-run `reconcile_tree()`. The three actions are:
- `"accept"` — confirm a specific `name_x → name_y` mapping.
- `"reject"` — mark a name as deliberately excluded.
- `"replace"` — redirect `name_x` to a different `name_y` than the
  cascade produced.

### Produce aligned objects

Once satisfied with the reconciliation, apply it:

```{r apply}
aligned <- reconcile_apply(
  result,
  data = trait_data,
  tree = tree,
  species_col = "species",
  drop_unresolved = TRUE
)

# Aligned data frame — only species present in both data and tree
aligned$data

# Aligned tree — pruned to matched species
ape::Ntip(aligned$tree)
```

The `$data` and `$tree` components now have matching species, ready for
comparative analysis.

## Example 2: Reconcile two datasets

When merging trait data from different sources, species names often
disagree:

```{r data-data}
df1 <- data.frame(
  species = c("Homo sapiens", "Pan_troglodytes", "Gorilla gorilla"),
  mass = c(70, 50, 160)
)

df2 <- data.frame(
  species = c("Homo sapiens", "Pan troglodytes", "Pongo pygmaeus"),
  lifespan = c(79, 40, 45)
)

result2 <- reconcile_data(
  x = df1,
  y = df2,
  authority = NULL,
  quiet = TRUE
)

print(result2)
```

`Pan_troglodytes` (underscore) in `df1` is matched to `Pan troglodytes`
(space) in `df2` via normalisation.

## Understanding match types

Every row in `reconcile_mapping()` has a `match_type` column. Here is
what each value means and what action (if any) it requires:

| `match_type`  | Meaning | Action needed? |
|---------------|---------|----------------|
| `exact`       | Verbatim string equality | None |
| `normalized`  | Names matched after stripping underscores, authority strings, and case differences | None — check the `notes` column if you want to confirm |
| `synonym`     | Names resolved through a taxonomic authority (e.g., Catalogue of Life) to the same accepted name | Verify the resolved name looks correct |
| `fuzzy`       | High-confidence character-level match (score ≥ `flag_threshold`, default 0.95) | Check the `match_score` column; review with `reconcile_suggest()` |
| `flagged`     | Lower-confidence match that needs human review: fuzzy score below `flag_threshold`, or an indirect synonym chain | **Review with `reconcile_review()` or `reconcile_suggest()`** |
| `manual`      | Set by `reconcile_override()` or the `overrides` argument | None — you decided this |
| `unresolved`  | No match found after all stages | Investigate; use `reconcile_suggest()` for candidates or `reconcile_override()` to document a decision |

Use `reconcile_summary(result, detail = "mismatches_only")` to see only
the rows that need attention.

## Example 3: Using a taxonomic authority

**When should you set `authority`?** Use `authority = NULL` (skip
synonym lookup) when:
- You want a quick offline check — no database download required.
- Your data and tree share the same taxonomy (most formatting differences
  are caught by the normalisation stage anyway).

Set `authority = "col"` (or another provider) when names differ because
of genuine taxonomic revisions — species moved to a different genus,
splits, or lumps. The first run will download a local database (~100 MB);
subsequent runs are fast because the database is cached.

When names differ due to synonymy, pass an authority to enable synonym
resolution via taxadb:

```{r authority, eval = FALSE}
# Requires taxadb and a local database download (automatic on first use)
result3 <- reconcile_tree(
  x = trait_data,
  tree = tree,
  x_species = "species",
  authority = "col"        # Catalogue of Life
)
```

Supported authorities: `col`, `itis`, `gbif`, `ncbi`, `ott`, and
`itis_test` (a small bundled subset of ITIS used for examples). These
mirror the providers documented in `?taxadb::td_create`; the database
is downloaded once on first use and cached locally, ensuring
reproducibility (results do not depend on an external API being
available).

## Example 4: Pre-built overrides

Labs often maintain a curated list of known corrections. You can pass
these as a data frame or CSV path:

```{r overrides-table, eval = FALSE}
# A data frame of known corrections
corrections <- data.frame(
  name_x = c("Corvus sp.", "Turdus merulaa"),
  name_y = c("Corvus corax", "Turdus merula"),
  user_note = c("Only one Corvus in our tree", "Typo in source data")
)

result4 <- reconcile_tree(
  x = my_data,
  tree = my_tree,
  overrides = corrections
)

# Or from a CSV file:
result5 <- reconcile_tree(
  x = my_data,
  tree = my_tree,
  overrides = "lab_corrections.csv"
)
```

Overrides are applied before any other matching stage, so they always
take priority.

## Example 5: Multiple datasets against one tree

`reconcile_multi()` reconciles several datasets at once, pooling all
unique species names before running the cascade:

```{r multi, eval = FALSE}
# Suppose you have several data frames to reconcile against one tree:
datasets <- list(
  traits  = trait_data,        # defined above
  ecology = my_ecology_data,   # your own data frame
  morpho  = my_morpho_data     # your own data frame
)

result6 <- reconcile_multi(datasets, my_tree)
print(result6)
```

## Key design principles

1. **Conservative**: Names are never silently changed. Ambiguous cases
   are flagged, not auto-resolved.
2. **Transparent**: Every decision is recorded with match type, score,
   source, and a human-readable note.
3. **Reproducible**: Database versions are pinned. All parameters are
   stored in the reconciliation object.
4. **Practical**: Works with the objects comparative biologists already
   use — `data.frame` and `ape::phylo`.

## Typical workflow

```{r workflow, eval = FALSE}
library(prepR4pcm)

# 1. Load your data and tree
my_data <- read.csv("species_traits.csv")
my_tree <- ape::read.tree("species_tree.nwk")

# 2. Reconcile
result <- reconcile_tree(my_data, my_tree, authority = "col")

# 3. Review
print(result)
reconcile_summary(result, detail = "mismatches_only")

# 4. Fix manually if needed
result <- reconcile_override(result, "Corvus sp.", "Corvus corax",
                             note = "Only one Corvus in tree")

# 5. Apply
aligned <- reconcile_apply(result, data = my_data, tree = my_tree,
                            drop_unresolved = TRUE)

# 6. Analyse
# aligned$data and aligned$tree are ready for caper, phytools, MCMCglmm, etc.
```

## References

- Hadfield, J.D. (2010) MCMC methods for multi-response generalized
  linear mixed models: the MCMCglmm R package. *Journal of Statistical
  Software* 33:1--22. [doi:10.18637/jss.v033.i02](https://doi.org/10.18637/jss.v033.i02)
- Orme, D., Freckleton, R., Thomas, G., Petzoldt, T., Fritz, S.,
  Isaac, N. & Pearse, W. (2025) caper: Comparative Analyses of
  Phylogenetics and Evolution in R. R package version 1.0.4.
  [doi:10.32614/CRAN.package.caper](https://doi.org/10.32614/CRAN.package.caper)
- Paradis, E. & Schliep, K. (2019) ape 5.0: an environment for modern
  phylogenetics and evolutionary analyses in R. *Bioinformatics*
  35:526--528. [doi:10.1093/bioinformatics/bty633](https://doi.org/10.1093/bioinformatics/bty633)
- Revell, L.J. (2024) phytools 2.0: an updated R ecosystem for
  phylogenetic comparative methods (and other things).
  *PeerJ* 12:e16505. [doi:10.7717/peerj.16505](https://doi.org/10.7717/peerj.16505)
