---
title: "Phylogenetic meta-analysis with rotl + prepR4pcm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic meta-analysis with rotl + prepR4pcm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # most chunks need rotl + metafor + network
)
```

Phylogenetic meta-analysis is the dominant comparative workflow in
ecology and evolutionary biology when the dataset spans many higher
taxa: you compute an effect size per study, then fit a model that
controls for shared evolutionary history. The standard recipe
(Cinar et al. 2022, *Methods Ecol. Evol.* 13:383) looks like:

1. **Pull a topology from Open Tree of Life** via `rotl`. OToL gives
   you tip labels and branching structure; the edge lengths it
   returns are unit-length placeholders — **not biologically
   meaningful** time/divergence values.
2. **Resolve polytomies at random** via `ape::multi2di(random =
   TRUE)`. The phylogenetic correlation matrix you'll feed into the
   meta-analysis model has to be full-rank, which means strictly
   bifurcating.
3. **Assign branch lengths via Grafen's (1989) method**:
   `ape::compute.brlen(tree, method = "Grafen")`. Grafen's method is
   the canonical choice for meta-analysis when no time-calibrated
   branches are available.
4. **Compute the phylogenetic correlation matrix**:
   `ape::vcv(tree, corr = TRUE)`.
5. **Fit the meta-analysis** with `metafor::rma.mv()` (or
   `MCMCglmm`, `brms`, `glmmTMB`, etc.) using the correlation matrix
   as a random-effect structure.

`prepR4pcm` wraps steps 1-3 into a one-call form, and step 4 into a
named helper. This vignette walks through the whole pipeline.

## A worked example: thermal-tolerance plasticity across animal classes

We'll mirror the phylogeny construction from Pottier et al. (2022)
"Developmental plasticity in thermal tolerance" (*Ecology
Letters* 25(10):2245-2268,
[doi:10.1111/ele.14083](https://doi.org/10.1111/ele.14083)). Their
dataset spans **147 ectotherm species across 14 animal classes** —
fish, insects, reptiles, amphibians, crustaceans, bivalves,
echinoderms, cephalopods, etc. This is exactly the cross-taxon
case where rotl is the only practical option (no taxon-specific
mega-tree covers all 14 classes).

For brevity, we'll use 13 representative species — one or two from
each major class in the Pottier dataset:

## Setup

```{r setup, eval = TRUE}
library(prepR4pcm)
# Suggested for the analysis itself:
# install.packages("rotl")     # OToL access
# install.packages("metafor")  # the meta-analysis engine
```

## Step 1-3: topology -> bifurcating -> Grafen branches, in one call

```{r get-tree}
species <- c(
  # Actinopterygii (ray-finned fish)
  "Salmo trutta",              "Acipenser fulvescens",
  # Reptilia
  "Anolis sagrei",
  # Amphibia
  "Xenopus laevis",            "Rhinella marina",
  # Insecta
  "Ischnura elegans",          "Anopheles albimanus",
  # Malacostraca (crustaceans)
  "Litopenaeus vannamei",
  # Branchiopoda
  "Daphnia magna",
  # Bivalvia (mussels)
  "Lampsilis abrupta",
  # Echinoidea (sea urchins)
  "Mesocentrotus franciscanus",
  # Cephalopoda
  "Octopus maya",
  # Chondrichthyes (sharks/rays)
  "Heterodontus portusjacksoni",
  # Hyperoartia (lampreys)
  "Petromyzon marinus"
)

res <- pr_get_tree(
  species,
  source             = "rotl",
  resolve_polytomies = TRUE,        # multi2di(random = TRUE)
  branch_lengths     = "grafen"     # compute.brlen(method = "Grafen")
)
res$tree           # ultrametric, bifurcating, Grafen branches
res$matched        # species OToL successfully resolved
res$unmatched      # species OToL couldn't place
```

The Pottier et al. methods description matches this almost exactly;
they wrote the pipeline out in `rotl::tnrs_match_names()` →
`rotl::tol_induced_subtree()` → `ape::multi2di(random = TRUE)` →
`ape::compute.brlen(method = "Grafen", power = 1)` →
`ape::vcv(cor = TRUE)`. The two new arguments to `pr_get_tree()`
collapse those four lines (steps 2-4) into one.

The two new arguments — `resolve_polytomies` and `branch_lengths` —
collapse what would otherwise be three explicit pipeline steps:

```{r equiv, eval = FALSE}
# What `pr_get_tree(... resolve_polytomies = TRUE,
#                       branch_lengths   = "grafen")`
# is equivalent to (when source = "rotl"):
res <- pr_get_tree(species, source = "rotl")  # rotl topology
tr  <- res$tree
tr  <- ape::multi2di(tr, random = TRUE)        # bifurcate
tr  <- ape::compute.brlen(tr, method = "Grafen")
res$tree <- tr
```

## Step 4: phylogenetic correlation matrix

```{r phylo-cor}
phy_cor <- pr_phylo_cor(res)
dim(phy_cor)            # 9 x 9 (one row/col per species)
isSymmetric(phy_cor)    # TRUE
all(diag(phy_cor) == 1) # TRUE (correlation matrix; ultrametric tree)
```

`pr_phylo_cor()` is a thin wrapper around `ape::vcv(tree, corr =
TRUE)`. The function exists separately so the meta-analysis pattern
is self-documenting and so the matrix can carry tree provenance.

## Step 5: fit the meta-analysis with `metafor`

For the Pottier-style analysis, the effect size per study is the
log response ratio (lnRR) of CTmax under contrasting developmental
temperatures. Sampling variance comes with each effect via the
group sample sizes / SDs. Suppose you've already extracted those
per-study, per-species (this is the realistic shape of an
ectotherm-thermal-tolerance dataset):

```{r meta}
library(metafor)

# Toy effect-size data --- one row per study; one or more rows per
# species. Replace with your real extraction.
set.seed(42)
toy_df <- data.frame(
  species = sample(species, 30, replace = TRUE),
  study   = sprintf("study_%02d", seq_len(30)),
  yi      = rnorm(30, mean = 0.20, sd = 0.15),   # lnRR
  vi      = runif(30, min = 0.01, max = 0.08)    # sampling variance
)

# Fit: random study effect (within-study) + random species effect
# (between-species, with phylogenetic correlation structure).
fit <- rma.mv(
  yi ~ 1,
  vi,
  random  = list(~1 | study, ~1 | species),
  R       = list(species = phy_cor),
  data    = toy_df,
  method  = "REML"
)
summary(fit)
```

The `R = list(species = phy_cor)` argument is what tells `metafor`
that the species random effect has the correlation structure given
by your phylogeny. The `~1 | study` partitions within-study from
between-species variance — standard for meta-analysis of
multi-study effect-size data.

Pottier et al. used a slightly more complex `metafor` model
including population, family, and shared-trial random effects (see
their `Statistical_analyses.Rmd`). The phylogenetic component is
the same `R = list(species = phy_cor)` either way.

## Manual additions when OToL doesn't have a species

A common situation: OToL doesn't have one of your species. Pottier
et al. ran into this with *Potamilus alatus* (a freshwater mussel)
— OToL's induced subtree errored on it, so they pulled it out of
the tnrs match, generated the tree without it, then grafted it
back manually with `ape::bind.tip()`:

```r
# Pottier et al.'s approach (paraphrased from their Rmd)
taxa.tree <- filter(taxa, ott_id != "732215") # remove Potamilus alatus
phylo_tree <- tol_induced_subtree(ott_ids = taxa.tree$ott_id)
binary.tree <- multi2di(phylo_tree, random = TRUE)
binary.tree <- bind.tip(binary.tree, tip.label = "Potamilus_alatus",
                        where = 95)            # grafted into Bivalvia
```

In `prepR4pcm`, the equivalent of "graft species back in" is
`reconcile_augment()`. After `pr_get_tree()` gives you the
OToL-resolved tree, call `reconcile_augment()` with the residually
unresolved species; it does the genus-level (or MRCA-level)
placement automatically and records the placement in `$augmented`
for the methods section:

```{r villosa, eval = FALSE}
# Step 0 in your workflow: reconcile data names against the tree.
# Any species OToL didn't place lands in `unresolved`; you can
# graft them onto the tree as sister to a congener (genus-level
# placement; cite the taxonomic source for the placement, just as
# Cinar et al. cite ITIS).
rec <- reconcile_data(my_effect_size_df, x_species = "species",
                       authority = "itis")
aug <- reconcile_augment(rec, res$tree)   # graft unresolved species
phy_cor <- pr_phylo_cor(aug$tree)
```

The same one-species-manually-added pattern shows up in Cinar et
al. (2022, *Methods Ecol. Evol.* 13:383), where *Villosa delumbis*
was added based on ITIS — they describe it as "Polytomies were
resolved at random, and one species, *Villosa delumbis*, was
manually added to the tree based on information from the
Integrated Taxonomic Information System."

## Why Grafen (and not datelife / a time-calibrated tree)?

In principle you could replace Grafen branches with biologically
real ones from datelife:

```{r datelife, eval = FALSE}
# Alternative: get rotl topology, then date it with datelife
res    <- pr_get_tree(species, source = "rotl",
                      resolve_polytomies = TRUE)   # NO branch_lengths
dated  <- pr_date_tree(res$tree)                   # datelife dates it
phy_cor <- pr_phylo_cor(dated)
```

For most phylogenetic meta-analyses, **we recommend Grafen**, not
datelife, for three reasons:

1. **Coverage.** Datelife synthesises from per-clade published
   chronograms. For a cross-taxon dataset like Pottier's (lampreys,
   bivalves, insects, fish, mammals etc., spanning 14 animal
   classes), datelife usually can't return a chronogram covering
   all of them — its database is sparse outside well-studied
   single-clade groups (mammals, birds, fish).
2. **Question alignment.** Phylogenetic meta-analysis models a
   *species random effect* with a phylogenetic correlation
   structure. The model asks "do related species have correlated
   effect sizes?", not "how fast does the trait evolve per unit
   time?". Pagel's λ on the species random effect (which `metafor`
   estimates implicitly when you supply the correlation matrix)
   absorbs branch-length scale, so the choice between Grafen-units
   and time-units matters less than it would for, say, a
   Brownian-motion rate estimate.
3. **Reproducibility.** Grafen's method is deterministic given the
   topology; datelife results depend on which chronograms are in
   their database the day you run it.

**Use datelife instead** when:

- Your taxa fall within a single, well-studied clade (mammals,
  birds, insects-Lepidoptera, fish — see `?pr_date_tree`).
- The science question requires real divergence times (trait-
  evolution rate, Brownian-motion variance per million years,
  divergence-time questions).
- You're fitting a Brownian-motion-on-tree model directly (e.g.
  in `caper::pgls` or `phylolm::phylolm`) where branch-length
  scale enters the variance directly.

For the realistic cross-taxon meta-analysis case (the one Pottier
et al. 2022 and many others run), Grafen via `pr_get_tree(...,
branch_lengths = "grafen")` is the practical default.

## What this vignette deliberately doesn't cover

- **Mixed-effect specification**. `metafor::rma.mv()` takes many more
  random/fixed structures than `~ 1 | species`. See Viechtbauer
  (2010) *Journal of Statistical Software* 36:1.
- **Branch-length sensitivity**. The Cinar et al. simulation shows
  Grafen-vs-time-calibrated lengths matter. If your topology *is*
  time-calibrated (e.g. `source = "fishtree"` or
  `source = "datelife"`), don't apply Grafen on top — pass
  `branch_lengths = NULL` (the default) and use the chronogram
  directly.
- **Multi-tree posterior**. If your backend returns a `multiPhylo`,
  `pr_phylo_cor()` returns a list of correlation matrices. Feed each
  to `metafor` separately and pool with Rubin's rules — see the
  [posterior-tree pipeline vignette](posterior-tree-pipeline.html).

## See also

- `?pr_get_tree` — the retrieval entry point.
- `?pr_phylo_cor` — the correlation-matrix wrapper.
- `?reconcile_augment` — for grafting species OToL didn't place.
- Pottier, Burke, Drobniak, Lagisz, Nakagawa (2022). *Developmental
  plasticity in thermal tolerance: ontogenetic variation, persistence,
  and future directions.* Ecology Letters 25(10): 2245--2268.
  [doi:10.1111/ele.14083](https://doi.org/10.1111/ele.14083).
  Source code: [github.com/p-pottier/Dev_plasticity_thermal_tolerance](https://github.com/p-pottier/Dev_plasticity_thermal_tolerance).
- Cinar, Nakagawa, & Viechtbauer (2022). *Phylogenetic multilevel
  meta-analysis: a simulation study on the importance of modelling
  the phylogeny.* Methods Ecol. Evol. 13(2): 383--395.
  [doi:10.1111/2041-210X.13760](https://doi.org/10.1111/2041-210X.13760).
- Michonneau, Brown, & Winter (2016). *rotl: an R package to
  interact with the Open Tree of Life data.* Methods Ecol. Evol.
  7(12): 1476--1481.
  [doi:10.1111/2041-210X.12593](https://doi.org/10.1111/2041-210X.12593).
- Paradis & Schliep (2019). *ape 5.0: an environment for modern
  phylogenetics and evolutionary analyses in R.* Bioinformatics
  35(3): 526--528.
  [doi:10.1093/bioinformatics/bty633](https://doi.org/10.1093/bioinformatics/bty633).
- Grafen (1989). *The phylogenetic regression.* Phil. Trans. R.
  Soc. B 326: 119--157.
  [doi:10.1098/rstb.1989.0106](https://doi.org/10.1098/rstb.1989.0106).
