---
title: "Phylogenetic meta-analysis with rotl + prepR4pcm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic meta-analysis with rotl + prepR4pcm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # most chunks need rotl + metafor + network
)
```

Phylogenetic meta-analysis is the dominant comparative workflow in
ecology and evolutionary biology when the dataset spans many higher
taxa: you compute an effect size per study, then fit a model that
controls for shared evolutionary history.

The standard recipe (Cinar et al. 2022, *Methods Ecol. Evol.*
13:383) looks like:

1. **Pull a topology from Open Tree of Life** via `rotl`. OTL gives
   you tip labels and branching structure; the edge lengths it
   returns are unit-length placeholders — **not biologically
   meaningful** time/divergence values.
2. **Resolve polytomies at random** via `ape::multi2di(random =
   TRUE)`. The phylogenetic correlation matrix you'll feed into the
   meta-analysis model has to be full-rank, which means strictly
   bifurcating.
3. **Assign branch lengths via Grafen's (1989) method**:
   `ape::compute.brlen(tree, method = "Grafen")`. Grafen's method is
   the canonical choice for meta-analysis when no time-calibrated
   branches are available.
4. **Compute the phylogenetic correlation matrix**:
   `ape::vcv(tree, corr = TRUE)`.
5. **Fit the meta-analysis** with `metafor::rma.mv()` (or
   `MCMCglmm`, `brms`, `glmmTMB`, etc.) using the correlation matrix
   as a random-effect structure.

`prepR4pcm` wraps steps 1-3 into a single `pr_get_tree()` call, and
step 4 into a single `pr_phylo_cor()` call. This vignette walks
through the whole pipeline.

## A worked example: thermal-tolerance plasticity across animal classes

We'll mirror the phylogeny construction from Pottier et al. (2022)
"Developmental plasticity in thermal tolerance" (*Ecology
Letters* 25(10):2245-2268,
[doi:10.1111/ele.14083](https://doi.org/10.1111/ele.14083)). Their
dataset spans **147 ectotherm species across 14 animal classes** —
fish, insects, reptiles, amphibians, crustaceans, bivalves,
echinoderms, cephalopods, etc.

This is the cross-taxon case where `rotl` is the only practical
option for retrieval. The taxon-specific mega-tree backends each
cover one clade only: `rtrees` ships separate mega-trees for
mammals, birds, fish, amphibians, squamates and a handful of other
taxa; `fishtree` covers ray-finned fish only; `clootl` covers birds
only. None of those backends spans the 14-class breadth of an
ectotherm meta-analysis. The Open Tree of Life synthesis tree that
`source = "rotl"` returns is the only single backend that places
species drawn from every class in one tree.

## Setup

We will use three packages: **prepR4pcm** (this package, for the
retrieve / clean / correlate pipeline), **rotl** (the Open Tree of
Life client that `pr_get_tree(source = "rotl")` calls under the
hood), and **metafor** (the meta-analysis engine that consumes the
phylogenetic correlation matrix).

```{r setup, eval = TRUE}
library(prepR4pcm)
# install.packages("rotl")     # required: OTL access (called by pr_get_tree)
# install.packages("metafor")  # required: the meta-analysis engine
```

## Step 1-3: topology -> bifurcating -> Grafen branches, in one call

For brevity we use **14 representative species** — one or two from
each major class in the Pottier dataset. Each class label below
points at one or two species drawn from a different chunk of the
ectotherm tree. We picked these classes because they are the ones
contributing the most studies in Pottier et al. (2022); a real
analysis would use all 147 species from the Pottier dataset.

```{r species-list}
species <- c(
  # Actinopterygii (ray-finned fish)
  "Salmo trutta",              "Acipenser fulvescens",
  # Reptilia
  "Anolis sagrei",
  # Amphibia
  "Xenopus laevis",            "Rhinella marina",
  # Insecta
  "Ischnura elegans",          "Anopheles albimanus",
  # Malacostraca (crustaceans)
  "Litopenaeus vannamei",
  # Branchiopoda
  "Daphnia magna",
  # Bivalvia (mussels)
  "Lampsilis abrupta",
  # Echinoidea (sea urchins)
  "Mesocentrotus franciscanus",
  # Cephalopoda
  "Octopus maya",
  # Chondrichthyes (sharks/rays)
  "Heterodontus portusjacksoni",
  # Hyperoartia (lampreys)
  "Petromyzon marinus"
)
length(species)   # 14
```

### The manual pipeline (without prepR4pcm)

Before introducing the wrapper, here is the **manual pipeline** the
Pottier et al. (2022) methods describe. It calls
`rotl::tnrs_match_names()` →
`rotl::tol_induced_subtree()` → `ape::multi2di(random = TRUE)` →
`ape::compute.brlen(method = "Grafen", power = 1)`. We use the
`_manual` suffix on every object so the wrapper version below can
sit next to it without overwriting anything:

```{r manual-pipeline, eval = FALSE}
# Step 1: get the rotl topology (no extras)
res_manual    <- pr_get_tree(species, source = "rotl")
tree_manual   <- res_manual$tree

# Step 2: resolve polytomies at random (bifurcating tree)
tree_manual   <- ape::multi2di(tree_manual, random = TRUE)

# Step 3: Grafen branch lengths
tree_manual   <- ape::compute.brlen(tree_manual, method = "Grafen")

res_manual$tree <- tree_manual
plot(tree_manual)
```

### The same in one call (with prepR4pcm)

`pr_get_tree()` with `source = "rotl"`, `resolve_polytomies = TRUE`,
and `branch_lengths = "grafen"` collapses the three steps above
into one call:

```{r get-tree}
res <- pr_get_tree(
  species,
  source             = "rotl",
  resolve_polytomies = TRUE,        # multi2di(random = TRUE)
  branch_lengths     = "grafen"     # compute.brlen(method = "Grafen")
)
res$tree            # ultrametric, bifurcating, Grafen branches
res$matched         # species OTL successfully resolved
res$unmatched       # species OTL couldn't place
plot(res$tree)      # visualise the topology
res$tree$tip.label  # has _ott<digits> suffixes (see note below)
```

**About the `_ott<digits>` suffixes.** Open Tree of Life identifies
every taxon by an OTT id (e.g. `Salmo_trutta_ott854188`). `rotl`
appends that id to the tip label so a downstream tool can verify
which OTT node was placed. The suffix is informative for
provenance but it makes the tip labels disagree with the binomial
names you almost certainly have in your trait data frame
(`"Salmo trutta"` vs `"Salmo_trutta_ott854188"`). The
meta-analysis step below strips the suffix and converts underscores
to spaces before computing the correlation matrix.

### Confirm the two pipelines produce the same tree

```{r compare-pipelines, eval = FALSE}
# Same tip set?
setequal(res$tree$tip.label, res_manual$tree$tip.label)   # TRUE
# Same branch-length range (modulo RNG order for the bifurcation)?
range(res$tree$edge.length)
range(res_manual$tree$edge.length)
# Side-by-side plot
op <- par(mfrow = c(1, 2))
plot(res_manual$tree, main = "manual ape pipeline")
plot(res$tree,        main = "one-call pr_get_tree()")
par(op)
```

## Step 4: phylogenetic correlation matrix

The tree currently has `_ott<digits>` suffixes and underscores in
its tip labels. Strip those before computing the correlation matrix
so the row / column names match the binomials in your trait data:

```{r clean-labels}
res$tree$tip.label <- gsub("_ott\\d+", "", res$tree$tip.label)
res$tree$tip.label <- gsub("_",        " ",  res$tree$tip.label)
res$tree$tip.label                          # now "Salmo trutta", etc.
```

Now compute the phylogenetic correlation matrix:

```{r phylo-cor}
phy_cor <- pr_phylo_cor(res)
dim(phy_cor)            # 14 x 14 (one row/col per species)
isSymmetric(phy_cor)    # TRUE
all(diag(phy_cor) == 1) # TRUE (correlation matrix; ultrametric tree)
rownames(phy_cor)       # matches the binomials in `species`
```

`pr_phylo_cor()` is a thin wrapper around `ape::vcv(tree, corr =
TRUE)`. The function exists separately so the meta-analysis pattern
is self-documenting and so the matrix can carry tree provenance.

## Step 5: fit the meta-analysis with `metafor`

For the Pottier-style analysis, the effect size per study is the
log response ratio (lnRR) of CTmax under contrasting developmental
temperatures. Sampling variance comes with each effect via the
group sample sizes / SDs. Suppose you've already extracted those
per-study, per-species (this is the realistic shape of an
ectotherm-thermal-tolerance dataset):

```{r meta}
library(metafor)

# Toy effect-size data. Start by including every species at least
# once (otherwise random sampling can leave some species out and
# rma.mv complains about "levels in species without matching rows
# in R"). Then add 16 extra studies sampled with replacement to get
# a realistic multi-study-per-species shape.
set.seed(42)
toy_df <- data.frame(
  species = c(species,                          # 14 guaranteed rows
              sample(species, 16, replace = TRUE)),  # 16 extra
  study   = sprintf("study_%02d", seq_len(30)),
  yi      = rnorm(30, mean = 0.20, sd = 0.15),   # lnRR
  vi      = runif(30, min = 0.01, max = 0.08)    # sampling variance
)

# Sanity-check: every species in the data is also a row in phy_cor.
stopifnot(all(toy_df$species %in% rownames(phy_cor)))
length(intersect(toy_df$species, rownames(phy_cor)))   # 14

# Fit: random study effect (within-study) + random species effect
# (between-species, with phylogenetic correlation structure).
fit <- rma.mv(
  yi ~ 1,
  vi,
  random  = list(~1 | study, ~1 | species),
  R       = list(species = phy_cor),
  data    = toy_df,
  method  = "REML"
)
summary(fit)
```

**Alternative: align tree and data via `reconcile_*` first.** If
your trait data has species that aren't in the tree (typos,
synonyms, OTL gaps), the manual cleanup above won't catch them.
A more robust pattern uses `reconcile_tree()` + `reconcile_apply()`
on the cleaned tree to drop unresolved species and produce a pair
guaranteed to match:

```{r reconcile-then-fit, eval = FALSE}
result <- reconcile_tree(
  x         = toy_df,
  tree      = res$tree,    # already gsub-cleaned, see Step 4
  x_species = "species",
  authority = NULL          # skip synonym lookup; tree labels already binomials
)
aligned <- reconcile_apply(
  result,
  data            = toy_df,
  tree            = res$tree,
  species_col     = "species",
  drop_unresolved = TRUE
)
phy_cor <- pr_phylo_cor(aligned$tree)
fit <- rma.mv(yi ~ 1, vi,
              random = list(~1 | study, ~1 | species),
              R      = list(species = phy_cor),
              data   = aligned$data,
              method = "REML")
```

The `R = list(species = phy_cor)` argument is what tells `metafor`
that the species random effect has the correlation structure given
by your phylogeny. The `~1 | study` partitions within-study from
between-species variance — standard for meta-analysis of
multi-study effect-size data.

Pottier et al. used a slightly more complex `metafor` model
including population, family, and shared-trial random effects (see
their `Statistical_analyses.Rmd`). The phylogenetic component is
the same `R = list(species = phy_cor)` either way.

## Manual additions when OTL doesn't have a species

A common situation: OTL doesn't have one of your species. Pottier
et al. ran into this with *Potamilus alatus* (a freshwater mussel)
— OTL's induced subtree errored on it, so they pulled it out of
the tnrs match, generated the tree without it, then grafted it
back manually with `ape::bind.tip()`:

```r
# Pottier et al.'s approach (paraphrased from their Rmd)
taxa.tree <- filter(taxa, ott_id != "732215") # remove Potamilus alatus
phylo_tree <- tol_induced_subtree(ott_ids = taxa.tree$ott_id)
binary.tree <- multi2di(phylo_tree, random = TRUE)
binary.tree <- bind.tip(binary.tree, tip.label = "Potamilus_alatus",
                        where = 95)            # grafted into Bivalvia
```

In `prepR4pcm`, the equivalent of "graft species back in" is
`reconcile_augment()`. After `pr_get_tree()` gives you the
OTL-resolved tree, call `reconcile_augment()` with the residually
unresolved species; it does the genus-level (or MRCA-level)
placement automatically and records the placement in `$augmented`
for the methods section:

```{r villosa, eval = FALSE}
# Step 0 in your workflow: reconcile data names against the tree.
# Any species OTL didn't place lands in `unresolved`; you can
# graft them onto the tree as sister to a congener (genus-level
# placement; cite the taxonomic source for the placement, just as
# Cinar et al. cite ITIS).
rec <- reconcile_data(my_effect_size_df, x_species = "species",
                       authority = "itis")
aug <- reconcile_augment(rec, res$tree)   # graft unresolved species
phy_cor <- pr_phylo_cor(aug$tree)
```

The same one-species-manually-added pattern shows up in Cinar et
al. (2022, *Methods Ecol. Evol.* 13:383), where *Villosa delumbis*
was added based on ITIS — they describe it as "Polytomies were
resolved at random, and one species, *Villosa delumbis*, was
manually added to the tree based on information from the
Integrated Taxonomic Information System."

## Why Grafen (and not datelife / a time-calibrated tree)?

In principle you could replace Grafen branches with biologically
real ones from datelife:

```{r datelife, eval = FALSE}
# Alternative: get rotl topology, then date it with datelife
res    <- pr_get_tree(species, source = "rotl",
                      resolve_polytomies = TRUE)   # NO branch_lengths
dated  <- pr_date_tree(res$tree)                   # datelife dates it
phy_cor <- pr_phylo_cor(dated)
```

For most phylogenetic meta-analyses, **we recommend Grafen**, not
datelife, for three reasons:

1. **Coverage.** DateLife synthesises from per-clade published
   chronograms. For a cross-taxon dataset like Pottier's (lampreys,
   bivalves, insects, fish, mammals etc., spanning 14 animal
   classes), datelife usually can't return a chronogram covering
   all of them — its database is sparse outside well-studied
   single-clade groups (mammals, birds, fish).
2. **Question alignment.** Phylogenetic meta-analysis models a
   *species random effect* with a phylogenetic correlation
   structure. The model asks "do related species have correlated
   effect sizes?", not "how fast does the trait evolve per unit
   time?". Pagel's λ on the species random effect (which `metafor`
   estimates implicitly when you supply the correlation matrix)
   absorbs branch-length scale, so the choice between Grafen-units
   and time-units matters less than it would for, say, a
   Brownian-motion rate estimate.
3. **Reproducibility.** Grafen's method is deterministic given the
   topology; datelife results depend on which chronograms are in
   their database the day you run it.

**Use datelife instead** when:

- Your taxa fall within a single, well-studied clade (mammals,
  birds, insects-Lepidoptera, fish — see `?pr_date_tree`).
- The science question requires real divergence times (trait-
  evolution rate, Brownian-motion variance per million years,
  divergence-time questions).
- You're fitting a Brownian-motion-on-tree model directly (e.g.
  in `caper::pgls` or `phylolm::phylolm`) where branch-length
  scale enters the variance directly.

For the realistic cross-taxon meta-analysis case (the one Pottier
et al. 2022 and many others run), Grafen via `pr_get_tree(...,
branch_lengths = "grafen")` is the practical default.

## What this vignette deliberately doesn't cover

- **Mixed-effect specification**. `metafor::rma.mv()` takes many more
  random/fixed structures than `~ 1 | species`. See Viechtbauer
  (2010) *Journal of Statistical Software* 36:1.
- **Branch-length sensitivity**. The Cinar et al. simulation shows
  Grafen-vs-time-calibrated lengths matter. If your topology *is*
  time-calibrated (e.g. `source = "fishtree"` or
  `source = "datelife"`), don't apply Grafen on top — pass
  `branch_lengths = NULL` (the default) and use the chronogram
  directly.
- **Multi-tree posterior**. If your backend returns a `multiPhylo`,
  `pr_phylo_cor()` returns a list of correlation matrices. Feed each
  to `metafor` separately and pool with Rubin's rules — see the
  [posterior-tree pipeline vignette](posterior-tree-pipeline.html).

## See also

- `?pr_get_tree` — the retrieval entry point.
- `?pr_phylo_cor` — the correlation-matrix wrapper.
- `?reconcile_augment` — for grafting species OTL didn't place.
- Pottier, Burke, Drobniak, Lagisz, Nakagawa (2022). *Developmental
  plasticity in thermal tolerance: ontogenetic variation, persistence,
  and future directions.* Ecology Letters 25(10): 2245--2268.
  [doi:10.1111/ele.14083](https://doi.org/10.1111/ele.14083).
  Source code: [github.com/p-pottier/Dev_plasticity_thermal_tolerance](https://github.com/p-pottier/Dev_plasticity_thermal_tolerance).
- Cinar, Nakagawa, & Viechtbauer (2022). *Phylogenetic multilevel
  meta-analysis: a simulation study on the importance of modelling
  the phylogeny.* Methods Ecol. Evol. 13(2): 383--395.
  [doi:10.1111/2041-210X.13760](https://doi.org/10.1111/2041-210X.13760).
- Michonneau, Brown, & Winter (2016). *rotl: an R package to
  interact with the Open Tree of Life data.* Methods Ecol. Evol.
  7(12): 1476--1481.
  [doi:10.1111/2041-210X.12593](https://doi.org/10.1111/2041-210X.12593).
- Paradis & Schliep (2019). *ape 5.0: an environment for modern
  phylogenetics and evolutionary analyses in R.* Bioinformatics
  35(3): 526--528.
  [doi:10.1093/bioinformatics/bty633](https://doi.org/10.1093/bioinformatics/bty633).
- Grafen (1989). *The phylogenetic regression.* Phil. Trans. R.
  Soc. B 326: 119--157.
  [doi:10.1098/rstb.1989.0106](https://doi.org/10.1098/rstb.1989.0106).
