---
title: "From species names to a phylogenetic posterior — prepR4pcm + pigauto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From species names to a phylogenetic posterior — prepR4pcm + pigauto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE      # most of this requires datelife / pigauto / network
)
```

Phylogenetic comparative methods (PCMs) propagate uncertainty
imperfectly. The standard pipeline takes one tree, fits one model,
and reports one set of confidence intervals. But the tree itself is
typically a point estimate from a posterior, the trait data
typically have missing values, and each of those is its own source
of uncertainty.

This vignette walks through a pipeline that handles both:

1. **prepR4pcm** reconciles species names, fetches a *posterior of
   trees* (not just one), and prepares the data for downstream
   modelling.
2. **[pigauto](https://itchyshin.github.io/pigauto/)** imputes the
   missing trait values, fits the model on each posterior tree, and
   pools the results via Rubin's rules so the reported standard
   errors reflect both the tree and imputation uncertainty.

The two packages compose: prepR4pcm produces `multiPhylo`, pigauto
consumes `multiPhylo`. Once the contract is set up, swapping the
backend (rtrees vs datelife vs fishtree) doesn't change the rest of
the pipeline.

## What we'll build

A model-ready dataset for fish, with:

- A posterior of 50 fish chronograms from the Fish Tree of Life
  (Rabosky et al. 2018).
- 5 imputations of the missing trait values per tree (so 5 × 50 =
  250 model fits — adjust `m_per_tree` upward for a real analysis).
- Pooled estimates with valid standard errors that account for both
  axes of uncertainty.

```{r setup}
library(prepR4pcm)
# Suggests packages we'll lean on:
# install.packages("fishtree")
# pak::pak("phylotastic/datelife")
# Optional, for the pooling step at the end:
# pak::pak("itchyshin/pigauto")
```

## Step 1 — Reconcile names

Start with whatever names appear in your dataset. They never quite
match the tree's tip labels.

```{r reconcile, eval = FALSE}
# Your data and your tree — typically loaded from disk
my_data <- data.frame(
  species   = c("Salmo salar", "Esox lucius", "Esox_lucius",
                "Oncorhynchus mykiss"),
  body_size = c(98, 130, NA, 50)   # NA is OK; pigauto will impute
)

# Reconcile names against a published taxonomic authority. Catches
# typos, formatting drift, and synonymy.
rec <- reconcile_data(my_data,
                      x_species = "species",
                      authority = "col")  # Catalogue of Life
rec   # one-line summary of matched / unmatched / counts
```

## Step 2 — Get a posterior of trees

Every backend that *can* return multiple trees does so via
`n_tree > 1`. Here we ask `fishtree` for 50 stochastic resolutions
(Chang, Rabosky, Alfaro 2019, *Syst. Biol.*).

```{r retrieve, eval = FALSE}
trees <- pr_get_tree(rec,
                     source = "fishtree",
                     n_tree = 50)

class(trees$tree)             # "multiPhylo"
length(trees$tree)            # 50
length(trees$matched)         # how many species placed
head(trees$unmatched)         # species absent from fishtree

# Per-tree provenance: one entry per tree
length(trees$backend_meta$tree_provenance)  # 50
trees$backend_meta$tree_provenance[[1]]
```

For the universal-but-slower option, use `source = "datelife"` —
the chronogram-database backend that returns one calibrated tree per
source (Hedges, Bininda-Emonds, Upham, etc.).

```{r datelife, eval = FALSE}
trees <- pr_get_tree(rec,
                     source = "datelife",
                     n_tree = 50)
# trees$tree is multiPhylo; per-source citations are in
# trees$backend_meta$tree_provenance[[i]]$citation
```

If you already have a topology and want to *date* it (rather than
fetch a new one), use `pr_date_tree()`:

```{r date, eval = FALSE}
my_topology <- ape::read.tree("my_topology.nex")
dated_trees <- pr_date_tree(my_topology, n_dated = 50)
class(dated_trees$tree)        # "multiPhylo"
length(dated_trees$tree)       # up to 50
```

**What `n_dated = 50` actually returns**: 50 chronograms that all
share `my_topology` but have *different branch lengths*. Each
variant comes from a different source paper in DateLife's database
(e.g. variant 1 dated against Hedges et al. 2015, variant 2 against
Bininda-Emonds et al. 2007, etc.). The topology is fixed; the dating
varies.

To get 50 different *topologies*, fetch them separately first (e.g.
`pr_get_tree(species, source = "rtrees", taxon = "mammal")` returns
100 mammal topologies sampled from VertLife / Upham et al. 2019),
then pass that `multiPhylo` to `pr_date_tree()` to date each one.
That gives you the topology × source cross-product.

## Step 3 — Format citations (do this *before* you submit)

Reviewers ask. Save yourself the round-trip.

```{r cite, eval = FALSE}
# Plain text for an email or a methods paragraph
pr_cite_tree(trees)

# Markdown for a GitHub issue or PR description
cat(pr_cite_tree(trees, format = "markdown"))

# BibTeX for a manuscript bibliography (sanity-check before submitting)
cat(pr_cite_tree(trees, format = "bibtex"))
```

## Step 4 — Hand the posterior to pigauto

This is the contract. `pigauto::multi_impute_trees(df, trees, m_per_tree)`
takes `df` (data with NAs) and `trees` (`multiPhylo`) and returns
`m_per_tree` complete-data imputations *per tree*. The output knows
which imputation came from which tree, and pigauto's pooling step
later uses that index.

```{r pigauto, eval = FALSE}
library(pigauto)

# 5 imputations per tree × 50 trees = 250 complete datasets
mi <- multi_impute_trees(rec$aligned,
                         trees      = trees$tree,
                         m_per_tree = 5)

# Fit your model to each completed dataset. Replace the formula
# with whatever your hypothesis is.
fits <- with_imputations(mi, function(df) {
  glmmTMB::glmmTMB(body_size ~ 1 + (1 | species),
                   data = df)
})

# Pool via Rubin's rules — the standard errors reflect BOTH the
# imputation uncertainty AND the tree-posterior uncertainty.
pooled <- pool_mi(fits,
                  coef_fun = function(fit) fixef(fit)$cond,
                  vcov_fun = function(fit) vcov(fit)$cond)

pooled
```

The reported intervals are wider than the equivalent
single-tree-single-imputation pipeline — and *correctly* so. The
narrower intervals were always overconfident.

## Choosing a backend

Verified on a clean macOS R 4.4 install on 2026-05-01. See the
[Comparing tree backends](comparing-tree-backends.html) vignette
for the full status table.

| If your taxon is... | And you want... | Use | Status |
|---|---|---|---|
| Universal | A single best-guess tree | `pr_get_tree(source = "rotl")` | ✅ verified — returns the Open Tree of Life synthesis tree |
| Universal | A posterior of dated trees | `pr_get_tree(source = "datelife", n_tree = 50)` | likely ✅ — `datelife` is in `Enhances`; install separately with `pak::pak("phylotastic/datelife")` |
| Birds | A single tree, current Clements | `pr_get_tree(source = "clootl", n_tree = 1)` | ✅ verified — works out of the box (uses the v1.6/2025 taxonomy bundled with `clootl`) |
| Birds | A posterior of trees, current Clements | `pr_get_tree(source = "clootl", n_tree = 50)` | ⚠️ requires `clootl::get_avesdata_repo(".")` once before this works; without it `sampleTrees()` errors with `AvesData repo not found`. Capped at 100 upstream. |
| Fish | Time-calibrated, posterior | `pr_get_tree(source = "fishtree", n_tree = 50)` | ✅ verified — returns `multiPhylo[50]` |
| Bird, mammal, fish, amphibian, reptile, plant, shark, bee, butterfly | Taxon-specific mega-tree, possibly grafted | `pr_get_tree(source = "rtrees", taxon = "<group>")` | ✅ works; ⚠️ `n_tree` is **informational only** — `rtrees::get_tree()` has no `n_tree` argument, so the count is fixed by the chosen mega-tree (`taxon = "bird"` → 100, `taxon = "mammal"` → 100, `taxon = "fish"` → 1, etc.). |

For the dating-only case (you already have a topology):

| If your topology is... | Use | Status |
|---|---|---|
| Universal taxa | `pr_date_tree(my_tree, n_dated = 50)` | likely ✅ — untested in this run; same dependency story as `datelife` |

## What's *not* in this pipeline

- **Tree comparison across backends.** prepR4pcm provides
  `pr_tree_compare()` for quick pairwise Jaccard / Robinson-Foulds /
  branch-length comparisons (see the
  [comparing tree backends vignette](comparing-tree-backends.html)).
  For richer visual comparison use `phytools::cophyloplot()` or
  `phangorn::densiTree()`.
- **Trait imputation alone.** That's pigauto's `impute()` /
  `multi_impute()`. This vignette skips straight to the
  trees-aware variant because that's the integration point.

For repeated runs of the same retrieval (e.g. while iterating on a
manuscript), pass `cache = TRUE` to `pr_get_tree()`. The cache lives
under `tools::R_user_dir()` by default; redirect with
`pr_tree_cache_dir()` if you want it inside the project.

## See also

- `?pr_get_tree`, `?pr_date_tree`, `?pr_cite_tree`,
  `?reconcile_augment` --- the four entry points for prepR4pcm's
  tree-handling.
- [pigauto's full PCM workflow vignette](https://itchyshin.github.io/pigauto/articles/mixed-types.html)
  --- the downstream half, in detail.
- Sanchez Reyes et al. (2024). *DateLife: Leveraging databases and
  analytical tools to reveal the dated Tree of Life.* Systematic
  Biology 73(2): 470–485.
  [doi:10.1093/sysbio/syae015](https://doi.org/10.1093/sysbio/syae015).
- Rabosky et al. (2018). *An inverse latitudinal gradient in
  speciation rate for marine fishes.* Nature 559: 392–395.
  [doi:10.1038/s41586-018-0273-1](https://doi.org/10.1038/s41586-018-0273-1).
